---
title: "Synonymous / non-synonymous sites"
output:
  html_document:
    df_print: paged
---

Having found some parts of the voltage-gated sodium channel gene and a few hundred SNPs withing those contigs, it would be nice to know if we have any potential non-synonymous changes.

# Libraries

```{r warning=FALSE}
library(VariantAnnotation)
library(txdbmaker)
library(stringr)
library(dplyr)
library(purrr)
```

# Data files

```{r}
fragmentsFasta <- "sodiumChannelFragments.fasta"
fragmentsGFF <- "../../putativeChannelFragmentsShortReadAssembly/sodiumChannelFragments.gff"
fragmentsVCF <- "../poolFstat/includedSites.vcf"
```

# Read in the data

```{r}
snps <- readVcf(fragmentsVCF)
```

```{r}
dna.sequence <- FaFile(fragmentsFasta)
```

```{r}
gene.info <- makeTxDbFromGFF(fragmentsGFF)

```


# Analyze for synonymous / non-synonymous

```{r}
coding.info <- predictCoding(query = snps,
                             subject = gene.info,
                             seqSource = dna.sequence)
```

Due to a peculiarity of how the alt alleles are represented in the VCF file (e.g. A,*) there is an extra entry for most sites that is marked "not translated". Get rid of these. (NOT RUN, fixed later)

```{r}
#coding.info <- coding.info[which(mcols(coding.info)$CONSEQUENCE != "not translated")]
```


How many synonymous/non-synonymous do we have?

```{r}
table(mcols(coding.info)$CONSEQUENCE)

```

Interestingly, we also have one nonsense polymorphism that introduces a stop codon.

List out the non-synonymous and nonsense polymorphisms.

```{r}
names(coding.info[which(mcols(coding.info)$CONSEQUENCE == "nonsynonymous")])

```
```{r}
names(coding.info[which(mcols(coding.info)$CONSEQUENCE == "nonsense")])

```
Hmmm, there is something fishy here - two of our non-synonymous SNPs are at the same position. Checking the raw VCF (which was prepared for poolfstat) the site does appear to be trialleleic. Digging some more, I see what is happening here. In the VCF, where there is a third allele, the read count is generally 1 or 2. Since the VCf is not explicitly filtered for vallid alleles, a sequencing error can sneak in.

Solution: Filter out extra "alleles" that are not the seregating bases called by SNAPE-pooled, getting rid of likely sequencing erors and just retaining the bases that were called as SNPs

```{r}
snapeSNPsFile <- "../variantCalling/SNAPE/combined.out"
snape.snps <- read.table(snapeSNPsFile)
names(snape.snps) <- c("scaffold",
                       "pos",
                       "ref",
                       "ref.count",
                       "alt.count",
                       "ref.qual",
                       "alt.qual",
                       "top.two.nucs",
                       "1-p(0)",
                       "p(1)",
                       "frequency")
```

Set up a key based on scaffold and position


```{r}
snape.snps$key <- paste(snape.snps$scaffold, 
                        snape.snps$pos, 
                        sep = ":")

```
Filter the snape SNPs info to just the sites with entries in `coding.info` - requires some gymnastics to get the names of the `coding.info` entries without the underscore followed by segregating bases.

```{r}
tmp <- paste(str_split(names(coding.info), 
                       "_", 
                       simplify = T)[,1], 
             str_split(names(coding.info), 
                       "_", 
                       simplify = T)[,2], 
             sep = "_")

#snape.snps <- snape.snps[snape.snps$key %in% tmp,]


```

Next we need a `data.frame` with one row per entry in `coding.info` that contains the `varAllele` value, plus scaffold, position as a key to use as a left join with the snape snps. We also need to get the strand info so we can convert the variant.allele back to the same strans as the SNAPE info

```{r}
coding.info.variants <- data.frame(key = paste(str_split(names(coding.info), 
                                                         "_", 
                                                         simplify = T)[,1], 
                                               str_split(names(coding.info), 
                                                         "_", 
                                                         simplify = T)[,2], 
                                               sep = "_"),
                                   variant.allele = as.character(coding.info$varAllele),
                                   strand = as.character(coding.info@strand)
                                   )
```
Now left join so that entries in snape snps are replicated for contig.info entris at same scaffold and position

```{r}
combined.coding.info.snape.snps <- left_join(coding.info.variants, snape.snps, by = "key")

```
Wrinkle: coding.info$varAllele is complemented if strand is set to -

A custom function to return the complementary base. `Biostrings::complement()` does this but it won't work directly on strings and trying to round-trip is a pain.

```{r}
complement.base <- function(base){
  res <- case_match(base,
                    "T" ~ "A",
                    "A" ~ "T",
                    "G" ~ "C",
                    "C" ~ "G",
                    NA ~ "")
  if (is.na(res)){
    return("")
    }
  return(res)
  }
```

Complement base if variant is on the negative strand, add to the `data.frame`

```{r}
complemented.bases <- sapply(combined.coding.info.snape.snps$variant.allele, complement.base)
corrected.bases <- if_else(combined.coding.info.snape.snps$strand == "-",
                           complemented.bases,
                           combined.coding.info.snape.snps$variant.allele)
names(corrected.bases) <- NULL

combined.coding.info.snape.snps$corrected.variantAllele <- corrected.bases
```

Now we should be able to figure out if the corrected variant allele appears in the top 2 nucleotides reported by SNAPE at the site. If it does, we keep it as a valid SNP.

Note that if we use "" as the pattern for `str_detect` it will return `NA`, so we need to account for that

```{r warning=FALSE}
tmp <- str_detect(combined.coding.info.snape.snps$top.two.nucs,
                  combined.coding.info.snape.snps$corrected.variantAllele)

combined.coding.info.snape.snps$valid.snp <- if_else(is.na(tmp), 
                                                     FALSE, 
                                                     tmp)

```


Now, after much finagling, we have a vector that will let us filter our `coding.info` elements to only the SNPs identified by SNAPE

```{r}
coding.info.valid <- coding.info[combined.coding.info.snape.snps$valid.snp]

```

How many SNPs fall in coding regions?

```{r}
length(coding.info.valid)
```

Now to count up synonymous versus non-synonymous sites

```{r}
table(mcols(coding.info.valid)$CONSEQUENCE)

```

Get the nonsynonymous sites

```{r}
coding.info.nonysnonymous <- coding.info.valid[mcols(coding.info.valid)$CONSEQUENCE == "nonsynonymous"]

```
```{r}
coding.info.nonysnonymous
```


*We have kdr!!!*

From the notes I made when annotating the sodium channel gene fragments:

Multiple alignment of the protein translation of contig_322045 to the rice and palm weevil proteins PLUS the housefly protein (Q94615) shows that we do have the classic *kdr* mutation site (L1014F in housefly). The position of the codon is (-)428 - (-)425 and the assembled sequence encodes an L, as expected.

The non-synonymous polymorphisms on contig_95372 are toward the N-terminus of the protein.

## Non-synonymous SNP frequencies.

We want to know the frequency of the kdr mutation globally and at each sample location. Luckily, we already have these from SNAPE-pooled. We even already have the global data loaded and set up with a key.

First, get the contig:position part of the non-synonymous sites info so we can use it to lookup.

```{r}
non_synonymous.keys <- str_remove(names(coding.info.nonysnonymous), "_[A-Z]/[A-Z]")
```

Lets take a look at the global frequencies

```{r}
snape.snps[snape.snps$key %in% non_synonymous.keys,c(1, 2, 8, 11)]

```
So, the kdr allele is at fairly low frequency, about 0.04.

## Site specific frequencies

What about at individual sample locations. We have individual SNAPE output files for each site.

```{r}
site.snape.files <- list.files("../variantCalling/SNAPE/samples/", "*.out")
site.snape.files <- paste("../variantCalling/SNAPE/samples", site.snape.files, sep = "/")
site.snape.files
```

Need to do a bit of looping here to build up a big data.frame of SNAPE output plus extra columns with the contig:position key and the location,

```{r}
## first location

site.snape.results <- read.table(site.snape.files[1])

names(site.snape.results) <- c("scaffold",
                               "pos",
                               "ref",
                               "ref.count",
                               "alt.count",
                               "ref.qual",
                               "alt.qual",
                               "top.two.nucs",
                               "1-p(0)",
                               "p(1)",
                               "frequency")


site.snape.results$location <- rep(str_split_i(str_remove(site.snape.files[1],
                                                          "../variantCalling/SNAPE/samples/"), 
                                               "\\.", 
                                               1), 
                                   length(site.snape.results$scaffold))

site.snape.results$key <- paste(site.snape.results$scaffold, 
                                site.snape.results$pos, 
                                sep = ":")


## remaining locations

for (i in 2:length(site.snape.files)){
  tmp <- read.table(site.snape.files[i],
                    fill = T)
  
  names(tmp) <- c("scaffold",
                  "pos",
                  "ref",
                  "ref.count",
                  "alt.count",
                  "ref.qual",
                  "alt.qual",
                  "top.two.nucs",
                  "1-p(0)",
                  "p(1)",
                  "frequency")
  
  
  tmp$location <- rep(str_split_i(str_remove(site.snape.files[i],
                                             "../variantCalling/SNAPE/samples/"), 
                                  "\\.", 
                                  1), 
                      length(tmp$scaffold))
  
  tmp$key <- paste(tmp$scaffold, 
                   tmp$pos, 
                   sep = ":")
  
  site.snape.results <- rbind(site.snape.results,
                              tmp)

}

```

For figures, we re-code site names as 1 - 8, 1 being the most northerly site. Add this information to the
site results:

```{r}
site.info <- read.table("siteInfo.txt",
                        header = T)
#rename OldName col to location for easier left joins
names(site.info)[1] <- "location"
site.info

```
Add the site location info and site numbers for manuscript to the sample allele frequency results.


```{r}
site.snape.results <- left_join(site.snape.results, 
                                site.info, 
                                by = "location")

```


Get just the site frequencies for non-synonymous sites.

```{r}
non_synonymous.site.freqs <- subset(site.snape.results, key %in% non_synonymous.keys)

```

## Plot of kdr frequencies by site

Subset the kdr frequencies, sort so that sites are ordered north to south.

```{r}
kdr.freqs <- subset(non_synonymous.site.freqs, key == "contig_322045:428")
kdr.freqs <- kdr.freqs[order(kdr.freqs$MSName, decreasing = F),]
kdr.freqs
```

Simple barplot

```{r}
barplot(height = kdr.freqs$frequency,
        names.arg = kdr.freqs$MSName,
        col = "steelblue2",
        xlab = "site",
        ylab = "frequency",
        main = "kdr allele frequency")
dev.copy2pdf(file = "kdr.frequency.barplot.pdf")
```
Since sites are not equally spaced, plot frequency against latitude.

```{r}
plot(kdr.freqs$Latitude,
     kdr.freqs$frequency,
     pch = 21,
     bg = "red",
     xlab = "latitude",
     ylab = "frequency",
     main = "kdr allele frequency")

dev.copy2pdf(file = "kdr.frequency.latitude.pdf")
```


















