---
title: "Length distribution of RSSW cytoshome P450 sequences"
output: html_notebook
---

We have a set of 81 sequences that are likely cytochrome P450s. Let's see how many are close to what we expect for a full length CYP protein.

# Libraries

```{r}
library(seqinr)
library(dplyr)
```

# Initial analysis

## Load the sequences

```{r}
fastaFile <- "cytochromeP450.fasta"
seqs <- read.fasta(file = fastaFile,
                  seqtype = "AA")
```

## Get the lengths

```{r}
cyp.lengths <- unlist(lapply(seqs, length))
```

## Intital length plot

Note a close to full length CYP is generally in the range of 350 - 500 residues

```{r}
length.dist <- hist(cyp.lengths, 
                    breaks = 50)

```
So, we have quite a few below 350, which are likely partial sequences. Interstingly, we also have three sequences that are > 700 residues. This is much larger than we expect for a CYP and may be other proteins or fused gene models.

What are the outsized sequences?

```{r}
names(cyp.lengths[cyp.lengths > 700])

```

# Analysis of edited seqs

Corrected the issues with the over-long sequences.

## Read the data

```{r}
editedFastaFile <- "cytochromeP450_edited.fasta"
edited.seqs <- read.fasta(file = editedFastaFile,
                  seqtype = "AA")
```
## Get the lengths

```{r}
edited.cyp.lengths <- unlist(lapply(edited.seqs, length))
```

## Plot the lengths

```{r}
edited.length.dist <- hist(edited.cyp.lengths, 
                    breaks = 50,
                    main = "Cytochrome P450 length distribution",
                    xlab = "residues",
                    ylab = "count",
                    col = "darkolivegreen1")

```

## Get (near) full length CYPs
```{r}
write.table(data.frame(name = names(edited.cyp.lengths[edited.cyp.lengths >= 350])),
            file = "cytochromeP450_long.list",
            row.names = F,
            col.names = F,
            quote = F)

```

